---
title: "Periodontology 2000 - Adam's dispersal analysis script (Figure 3)" 
author: "Adam Burns"
date: "October 18, 2018"
output: html_document


---

```{r, echo=TRUE, error=FALSE, warning=FALSE, results=TRUE, message=FALSE}
###Load required libraries
library(knitr)
library(DESeq2)
library(phyloseq)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(vegan)
library(untb)
library(minpack.lm)
library(Hmisc)


###Import data
ddat.phy <- readRDS("~/Dropbox/oral-clinical-data-analysis-data/periodontology2000_hypo.RDS")
ddat.otu <- as.data.frame(otu_table(ddat.phy))
ddat.map <- as.data.frame(sample_data(ddat.phy))
ddat.tax <- as.data.frame(tax_table(ddat.phy))

#Calculate sampling depth
num.seqs <- rowSums(ddat.otu)
num.seqs <- as.data.frame(sort(num.seqs))
colnames(num.seqs) <- "num.seqs"
num.map <- merge(ddat.map, num.seqs, by.x=0, by.y=0, all=TRUE)
rownames(num.map) <- num.map$Row.names
num.map <- num.map[,-which(names(num.map) %in% c("Row.names"))]

###Sloan neutral community model
#Define neutral model fiting function
sncm.fit <- function(spp){
	require(minpack.lm)
	options(warn=-1)
	N <- mean(apply(spp, 1, sum))
	p <- apply(spp, 2, mean)/N
	spp.bi <- 1*(spp>0)
	freq <- apply(spp.bi, 2, mean)
	d <- 1/N
	m.fit <- nlsLM(freq ~ pbeta(d, N*m*p, N*m*(1-p), lower.tail=FALSE), start=list(m=0.1))
	m.ci <- confint(m.fit, 'm', level=0.95)
	#Calculate goodness-of-fit (R-squared and Root Mean Squared Error)
	freq.pred <- pbeta(d, N*coef(m.fit)*p, N*coef(m.fit)*(1-p), lower.tail=FALSE)
	Rsqr <- 1 - (sum((freq - freq.pred)^2))/(sum((freq - mean(freq))^2))
	RMSE <- sqrt(sum((freq-freq.pred)^2)/(length(freq)-1))
	#Results
	fitstats <- data.frame(m=numeric(), m.ci=numeric(), Rsqr=numeric(), RMSE=numeric())
	fitstats[1,] <- c(coef(m.fit), coef(m.fit)-m.ci[1], Rsqr, RMSE)
	return(fitstats)
}
#Define function for predicted values
sncm.pred <- function(spp){
	require(minpack.lm)
	require(Hmisc)
	options(warn=-1)
	N <- mean(apply(spp, 1, sum))
	p <- apply(spp, 2, mean)/N
	spp.bi <- 1*(spp>0)
	freq <- apply(spp.bi, 2, mean)
	d <- 1/N
	m.fit <- nlsLM(freq ~ pbeta(d, N*m*p, N*m*(1-p), lower.tail=FALSE), start=list(m=0.1))
	#Calculate predicted values
	freq.pred <- pbeta(d, N*coef(m.fit)*p, N*coef(m.fit)*(1-p), lower.tail=FALSE)
	#Calculate residuals
	pred.res <- freq - freq.pred
	#Results
	A <- cbind(p, freq, freq.pred, pred.res)
	A <- as.data.frame(A)
	colnames(A) <- c('p', 'freq', 'freq.pred', 'pred.res')
	return(A)
}

###Rarefaction analysis
#Select maximum sampling depth and remove samples under that depth
min.seq <- 50000
ddatr.otu <- ddat.otu[rowSums(ddat.otu) >= min.seq, ]
ddatr.otu <- rrarefy(ddatr.otu, min.seq)
ddatr.otu <- ddatr.otu[,(colSums(ddatr.otu) != 0)]
ddatr.map <- ddat.map[(rownames(ddat.map) %in% rownames(ddatr.otu)),]
#Define rarefaction levels to analyze
rare.lvls <- seq.int(from=1000, to=50000, by=1000)
#Fit model at different rarefaction depths
ncm.stats <- data.frame(sncm_m=numeric(), sncm_m.ci=numeric(), sncm_Rsqr=numeric(), sncm_RMSE=numeric(), Rarefaction=numeric())
for(i in 1:length(rare.lvls)){
	#Rarefy
	otu.i <- rrarefy(ddatr.otu, rare.lvls[i])
	otu.i <- otu.i[,(colSums(otu.i) != 0)]
	#Fit models
	A.i <- sncm.fit(otu.i)
	fit.stats.i <- c(A.i$m, A.i$m.ci, A.i$Rsqr, A.i$RMSE, rare.lvls[i])
	names(fit.stats.i) <- c('sncm_m', 'sncm_m.ci', 'sncm_Rsqr', 'sncm_RMSE', 'Rarefaction')
	ncm.stats[i,] <- fit.stats.i
}

#Rarefy to 10,000 sequences per sample
min.seq <- 10000
otu.r <- ddat.otu[rowSums(ddat.otu) >= min.seq, ]
otu.r <- rrarefy(otu.r, min.seq)
otu.r <- otu.r[,(colSums(otu.r) != 0)]
map.r <- ddat.map[(rownames(ddat.map) %in% rownames(otu.r)),]

#Fit model by habitat (gingival and tooth aspect) within subject
ncm.stats <- data.frame(Metacommunity=character(), Subject=character(), Aim=character(), Habitat_Class=character(), Tooth_Aspect=character(), sncm_m=numeric(), sncm_m.ci=numeric(), sncm_Rsqr=numeric(), sncm_RMSE=numeric(), No.Samples=numeric(), stringsAsFactors=FALSE)
for(s in 1:length(unique(map.r$Subject))){
	s.id <- unique(map.r$Subject)[s]
	map.s <- subset(map.r, Subject == s.id)
	aim.s <- unique(map.s$Aim)
	for(h in 1:length(unique(map.s$Habitat_Class))){
		habitat <- unique(map.s$Habitat_Class)[h]
		map.h <- subset(map.s, Habitat_Class == habitat)
		for(a in 1:length(unique(map.h$Tooth_Aspect))){
			aspect <- unique(map.h$Tooth_Aspect)[a]
			map.a <- subset(map.h, Tooth_Aspect == aspect)
			otu.a <- otu.r[(rownames(otu.r) %in% rownames(map.a)),]
			otu.a <- otu.a[,(colSums(otu.a) != 0)]
			fit.a <- sncm.fit(otu.a)
			meta.no <- (4*s) + (2*h) + (1*a) - 6 #Used the 'solve' function to determine coefficients
			vec.a <- c(as.character(paste('M',meta.no, sep='')), as.character(s.id), as.character(aim.s), as.character(habitat), as.character(aspect), fit.a$m, fit.a$m.ci, fit.a$Rsqr, fit.a$RMSE, nrow(map.a))
			names(vec.a) <- c('Metacommunity', 'Subject', 'Aim', 'Habitat_Class', 'Tooth_Aspect', 'sncm_m', 'sncm_m.ci', 'sncm_Rsqr', 'sncm_RMSE', 'No.Samples')
			ncm.stats[meta.no,] <- vec.a
		}
	}
}
ncm.stats$Aim <- factor(ncm.stats$Aim, levels=c('SA1', 'SA3'))
ncm.stats$Habitat_Class <- factor(ncm.stats$Habitat_Class, levels=c('Supra', 'Sub'))
ncm.stats$Tooth_Aspect <- factor(ncm.stats$Tooth_Aspect, levels=c('Buccal', 'Lingual'))

#Fit model by habitat (gingival and tooth aspect) within subject and get predictions
ncm.preds <- data.frame(OTU=character(), Metacommunity=character(), Subject=character(), Aim=character(), Habitat_Class=character(), Tooth_Aspect=character(), sncm_m=numeric(), sncm_m.ci=numeric(), sncm_Rsqr=numeric(), sncm_RMSE=numeric(), No.Samples=numeric(), p=numeric(), freq=numeric(), freq.pred=numeric(), pred.res=numeric(), stringsAsFactors=FALSE)
ncm.preds[1,] <- rep(NA, times=length(ncol(ncm.preds))) 
for(s in 1:length(unique(map.r$Subject))){
	s.id <- unique(map.r$Subject)[s]
	map.s <- subset(map.r, Subject == s.id)
	aim.s <- unique(map.s$Aim)
	for(h in 1:length(unique(map.s$Habitat_Class))){
		habitat <- unique(map.s$Habitat_Class)[h]
		map.h <- subset(map.s, Habitat_Class == habitat)
		for(a in 1:length(unique(map.h$Tooth_Aspect))){
			aspect <- unique(map.h$Tooth_Aspect)[a]
			map.a <- subset(map.h, Tooth_Aspect == aspect)
			otu.a <- otu.r[(rownames(otu.r) %in% rownames(map.a)),]
			otu.a <- otu.a[,(colSums(otu.a) != 0)]
			fit.a <- sncm.fit(otu.a)
			pred.a <- sncm.pred(otu.a)
			meta.no <- (4*s) + (2*h) + (1*a) - 6 #Used the 'solve' function to determine coefficients
			df.a <- cbind(rownames(pred.a), rep(as.character(paste('M',meta.no, sep='')), times=nrow(pred.a)), rep(as.character(s.id), times=nrow(pred.a)), rep(as.character(aim.s), times=nrow(pred.a)), rep(as.character(habitat), times=nrow(pred.a)), rep(as.character(aspect), times=nrow(pred.a)), rep(fit.a$m, times=nrow(pred.a)), rep(fit.a$m.ci, times=nrow(pred.a)), rep(fit.a$Rsqr, times=nrow(pred.a)), rep(fit.a$RMSE, times=nrow(pred.a)), rep(nrow(map.a), times=nrow(pred.a)), pred.a)
			colnames(df.a) <- c('OTU', 'Metacommunity', 'Subject', 'Aim', 'Habitat_Class', 'Tooth_Aspect', 'sncm_m', 'sncm_m.ci', 'sncm_Rsqr', 'sncm_RMSE', 'No.Samples', 'p', 'freq', 'freq.pred', 'pred.res')
			ncm.preds <- rbind(ncm.preds, df.a)
		}
	}
}
ncm.preds <- ncm.preds[-1,]
#Combine with taxonomic calls
ncm.preds <- merge(ncm.preds, ddat.tax, by.x=1, by.y=0, all=FALSE)
ncm.preds$Aim <- factor(ncm.preds$Aim, levels=c('SA1', 'SA3'))
ncm.preds$Habitat_Class <- factor(ncm.preds$Habitat_Class, levels=c('Supra', 'Sub'))
ncm.preds$Tooth_Aspect <- factor(ncm.preds$Tooth_Aspect, levels=c('Buccal', 'Lingual'))


#Variance stabilization via DeSeq2 (From McMurdie)
#Function for geometric mean, set to zero when all coordinates are zero
geo_mean_protected <- function(x) {
  if (all(x == 0)) {
    return (0)
  }
  exp(mean(log(x[x != 0])))
}
ps_dds <- phyloseq_to_deseq2(ddat.phy, design = ~ Habitat_Class)
geoMeans <- apply(counts(ps_dds), 1, geo_mean_protected)
ps_dds <- estimateSizeFactors(ps_dds, geoMeans = geoMeans)
ps_dds <- estimateDispersions(ps_dds)
ddat.otu.vs <- t(getVarianceStabilizedData(ps_dds))
ddat.otu.vs[ddat.otu.vs<0] <- 0 #Replace negative values with 0


#Estimate migration rates using Gst method by habitat (gingival and tooth aspect) within subject
#On variance stabilized data:
cols <- c(colnames(ddat.map), 'I', 'm', 'No.Samples')
gst.param <- t(as.data.frame(rep(NA, times=length(cols)), row.names=cols))
for(s in 1:length(unique(ddat.map$Subject))){
	s.id <- unique(ddat.map$Subject)[s]
	map.s <- subset(ddat.map, Subject == s.id)
	aim.s <- unique(map.s$Aim)
	for(h in 1:length(unique(map.s$Habitat_Class))){
		habitat <- unique(map.s$Habitat_Class)[h]
		map.h <- subset(map.s, Habitat_Class == habitat)
		for(a in 1:length(unique(map.h$Tooth_Aspect))){
			aspect <- unique(map.h$Tooth_Aspect)[a]
			map.a <- subset(map.h, Tooth_Aspect == aspect)
			otu.a <- ddat.otu.vs[(rownames(ddat.otu.vs) %in% rownames(map.a)),]
			otu.a <- otu.a[,(colSums(otu.a) != 0)]
			gst.a <- as.data.frame(optimal.params.gst(D=t(otu.a), exact=FALSE, ci=FALSE))
			gst.a <- cbind(gst.a, rep(nrow(map.a), times=nrow(map.a)))
			colnames(gst.a) <- c('I', 'm', 'No.Samples')
			df.a <- cbind(map.a, gst.a)
			gst.param <- rbind(gst.param, df.a)
		}
	}
}
gst.param <- gst.param[-1,]
gst.param$Aim <- factor(gst.param$Aim, levels=c('SA1', 'SA3'))
gst.param$Habitat_Class <- factor(gst.param$Habitat_Class, levels=c('Supra', 'Sub'))
gst.param$Tooth_Aspect <- factor(gst.param$Tooth_Aspect, levels=c('Buccal', 'Lingual'))

#On rarefied data:
cols <- c(colnames(map.r), 'I', 'm', 'No.Samples')
gst.param.r <- t(as.data.frame(rep(NA, times=length(cols)), row.names=cols))
for(s in 1:length(unique(map.r$Subject))){
	s.id <- unique(map.r$Subject)[s]
	map.s <- subset(map.r, Subject == s.id)
	aim.s <- unique(map.s$Aim)
	for(h in 1:length(unique(map.s$Habitat_Class))){
		habitat <- unique(map.s$Habitat_Class)[h]
		map.h <- subset(map.s, Habitat_Class == habitat)
		for(a in 1:length(unique(map.h$Tooth_Aspect))){
			aspect <- unique(map.h$Tooth_Aspect)[a]
			map.a <- subset(map.h, Tooth_Aspect == aspect)
			otu.a <- otu.r[(rownames(otu.r) %in% rownames(map.a)),]
			otu.a <- otu.a[,(colSums(otu.a) != 0)]
			gst.a <- as.data.frame(optimal.params.gst(D=t(otu.a), exact=FALSE, ci=FALSE))
			gst.a <- cbind(gst.a, rep(nrow(map.a), times=nrow(map.a)))
			colnames(gst.a) <- c('I', 'm', 'No.Samples')
			df.a <- cbind(map.a, gst.a)
			gst.param.r <- rbind(gst.param.r, df.a)
		}
	}
}
gst.param.r <- gst.param.r[-1,]
gst.param.r$Aim <- factor(gst.param.r$Aim, levels=c('SA1', 'SA3'))
gst.param.r$Habitat_Class <- factor(gst.param.r$Habitat_Class, levels=c('Supra', 'Sub'))
gst.param.r$Tooth_Aspect <- factor(gst.param.r$Tooth_Aspect, levels=c('Buccal', 'Lingual'))
```

# Part 1: Variation in microbial migration rates by habitat and disease state.
In order to determine whether microbial migration rates differed by tooth habitat, or by disease state, we fit the model separately to samples within each tooth habitat on an individual by individual basis.

### Figure 3A: Estimated migration rates of tooth habitats (Sloan model)
The estimated migration rates by habitat type and disease state. Each open point represents the estimated migration rate across samples for each habitat within each individual. Solid points represent the average across individuals for that habitat, while error bars represent the standard error.
```{r, echo=FALSE, error=FALSE, warning=FALSE}
##Plot (by individual - gingival and tooth aspect)
ncm.stats$Aim = factor(as.character(ncm.stats$Aim))
ncm.stats$Cohort = ifelse(ncm.stats$Aim == "SA1", "Control Cohort", "Low Flow Cohort")
p = ggplot(ncm.stats, aes(x=Tooth_Aspect, y=as.numeric(sncm_m))) +
	#geom_boxplot(aes(colour = Aim), outlier.shape=NA) +
	geom_point(aes(colour = Cohort), position=position_jitterdodge(jitter.width = 0, dodge.width = 0.8), shape=1, size=1, alpha=1) +
	stat_summary(aes(colour = Cohort), fun.data = mean_se, geom = "errorbar", position=position_dodge(0.8), width=0.1) +
	stat_summary(aes(colour = Cohort), fun.y = mean, geom = "point", position=position_dodge(0.8)) +
	facet_wrap(~Habitat_Class, strip.position="top", nrow=1) +
	scale_colour_manual(values = c('Control Cohort'='#377eb8', 'Low Flow Cohort'='#e41a1c')) +
	scale_y_log10() +
	ylab("Estimated migration rate (log(m))") +
	xlab("Tooth Habitat") +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   strip.background = element_blank(),
	   strip.text.x = element_text(colour='black'),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black', angle=30, vjust=1, hjust=1),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),,
	   legend.position='right',
	   legend.key = element_blank())
```


###Table XX: Analysis of variance for estimated migration rates across habitats and disease (Sloan model)
```{r, echo=FALSE, error=FALSE, warning=FALSE}
ncm.stats$Habitat_Class = ifelse(ncm.stats$Habitat_Class == "Sub","Subgingival", "Supragingival")
aov.fit <- with(ncm.stats, aov(as.numeric(as.character(sncm_m)) ~ Habitat_Class*Tooth_Aspect*Cohort))
summ.aov <- summary(aov.fit)[[1]]
kable(as.data.frame(summ.aov), caption="Analysis of variance on estimated migration rates")
```


###Figure 3A (v2): Estimated migration rates of tooth habitats (Gst)
```{r, echo=FALSE, error=FALSE, warning=FALSE}
##Plot (by individual - gingival and tooth aspect)
gst.param$Aim = factor(as.character(gst.param$Aim))
gst.param$Cohort = ifelse(gst.param$Aim == "SA1", "Control Cohort", "Low Flow Cohort")
gst.param$Habitat_Class = ifelse(gst.param$Habitat_Class=="Sub", "Subgingival", "Supragingival")
p = ggplot(gst.param, aes(x=Tooth_Aspect, y=as.numeric(as.character(m)))) +
	geom_boxplot(aes(colour = Cohort), outlier.shape=NA) +
	geom_point(aes(colour = Cohort), position=position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size=1, alpha=1) +
	#stat_summary(aes(colour = Cohort), fun.data = mean_se, geom = "errorbar", position=position_dodge(0.8), width=0.1) +
	#stat_summary(aes(colour = Cohort), fun.y = mean, geom = "point", position=position_dodge(0.8)) +
	facet_wrap(~Habitat_Class, strip.position="top", nrow=1) +
	scale_colour_manual(values = c('Control Cohort'='#377eb8', 'Low Flow Cohort'='#e41a1c')) +
	scale_y_log10() +
	ylab("Estimated immigration rate (log(m))") +
	xlab("Tooth Habitat") +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   strip.background = element_blank(),
	   strip.text.x = element_text(colour='black'),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black', angle=30, vjust=1, hjust=1),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),,
	   legend.position='right',
	   legend.key = element_blank())
```


###Table XX: Analysis of variance for estimated migration rates across habitats and disease (Gst)
```{r, echo=FALSE, error=FALSE, warning=FALSE}
aov.fit <- with(gst.param, aov(as.numeric(as.character(m)) ~ Habitat_Class*Tooth_Aspect*Cohort))
summ.aov <- summary(aov.fit)[[1]]
kable(as.data.frame(summ.aov), caption="Analysis of variance on estimated migration rates")
```



#Part 2: The aggregate fit of the neutral model by disease state and habitat
For visualization purposes, the following plots show the fit of the model averaged across individuals. Note, however, that the analyses in the test and other figures were done by fitting the model to groups of samples within individuals.

###Figure 3B-E: Observed and predicted distribution of microbial taxa
The fit of the neutral model to observed data averaged across individuals by habitat and disease state. The black points represent individual taxa, while the blue line represents the predicted values.
```{r, echo=FALSE, error=FALSE, warning=FALSE, results=FALSE, message=FALSE}
tax.focus <- c('SR1_3', 'SR1_2', 'Treponema_sp.1', 'Porphyromonas_gingivalis', 'Tannerella_sp.1', 'Tannerella_forsythia', 'Rothia_dentocariosa', 'Streptococcus_sanguinis', 'Abiotrophia_defectiva', 'Scardovia_wiggsiae', 'Streptococcus_mutans', 'Treponema_denticola')
tax.focus <- data.frame(cbind(tax.focus, seq(from = 97, to = 96+length(tax.focus))))
colnames(tax.focus) <- c('OTU', 'code')

#Healthy Supragingival
dat <- subset(ncm.preds, Aim=='SA1' & Habitat_Class=='Supra')
B.1 <- as.data.frame(cbind(with(dat, aggregate(p, list(OTU), mean))[,2], with(dat, aggregate(freq, list(OTU), mean))[,2], with(dat, aggregate(OTU, list(OTU), function(x){as.character(unique(x))}))[,2]))
colnames(B.1) <- c('p.mean', 'freq.mean', 'OTU')
B.1$p.mean <- as.numeric(as.character(B.1$p.mean))
B.1$freq.mean <- as.numeric(as.character(B.1$freq.mean))
tax = data.frame(ddat.phy@tax_table@.Data)
tax$OTU = rownames(tax)
tax$OTU = as.factor(as.character(tax$OTU))




Bf.1 <- merge(B.1, tax)
Bf.1$cat = ifelse(Bf.1$p.mean < 0.01, "low", "high")
m.fit <- with(B.1, nlsLM(freq.mean ~ pbeta(1/min.seq, min.seq*m*p.mean, min.seq*m*(1-p.mean), lower.tail=FALSE), start=list(m=0.1)))
fun.pred.1 <- function(x){pbeta(1/min.seq, min.seq*coef(m.fit)*x, min.seq*coef(m.fit)*(1-x), lower.tail=FALSE)}
plot.sa1_supra <- ggplot(B.1, aes(x=p.mean, y=freq.mean)) +
	geom_point() +
	geom_point(aes(x=p.mean, y=freq.mean), data=Bf.1, colour='white') +
	geom_point(aes(x=p.mean, y=freq.mean, color=Genus, shape=cat), data=Bf.1) +
	stat_function(data=B.1, aes(x=p.mean, y=freq.mean), fun=fun.pred.1, colour='#0072B2') +
	scale_x_log10() +
	ylab("Occurrence frequency") +
	xlab("log(Mean Relative Abundance)") +
	ggtitle("Supragingival communities across control subjects") +
	coord_cartesian(ylim=c(-0.025,1.025)) +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),
	   legend.position='none')

plot.sa1_supra
```


```{r}

#Healthy Subgingival
dat <- subset(ncm.preds, Aim=='SA1' & Habitat_Class=='Sub')
B.2 <- as.data.frame(cbind(with(dat, aggregate(p, list(OTU), mean))[,2], with(dat, aggregate(freq, list(OTU), mean))[,2], with(dat, aggregate(OTU, list(OTU), function(x){as.character(unique(x))}))[,2]))
colnames(B.2) <- c('p.mean', 'freq.mean', 'OTU')
B.2$p.mean <- as.numeric(as.character(B.2$p.mean))
B.2$freq.mean <- as.numeric(as.character(B.2$freq.mean))
Bf.2 <- subset(B.2, OTU %in% tax.focus$OTU)
Bf.2 <- merge(Bf.2, tax.focus, by.x='OTU', by.y='OTU', all=FALSE)
m.fit <- with(B.2, nlsLM(freq.mean ~ pbeta(1/min.seq, min.seq*m*p.mean, min.seq*m*(1-p.mean), lower.tail=FALSE), start=list(m=0.1)))
fun.pred.2 <- function(x){pbeta(1/min.seq, min.seq*coef(m.fit)*x, min.seq*coef(m.fit)*(1-x), lower.tail=FALSE)}

plot.sa1_sub <- ggplot(B.2, aes(x=p.mean, y=freq.mean)) +
	geom_point() +
	geom_point(aes(x=p.mean, y=freq.mean), data=Bf.2, colour='white') +
	geom_point(aes(x=p.mean, y=freq.mean, shape=code), data=Bf.2, colour='red', size=4) +
	scale_shape_manual(values=as.numeric(as.character(Bf.2$code))) +
	stat_function(data=B.2, aes(x=p.mean, y=freq.mean), fun=fun.pred.2, colour='#0072B2') +
	scale_x_log10() +
	ylab("Occurrence frequency") +
	xlab("log(Mean Relative Abundance)") +
	ggtitle("Subgingival communities across control subjects") +
	coord_cartesian(ylim=c(-0.025,1.025)) +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),
	   legend.position='none')


#Disease Supragingival
dat <- subset(ncm.preds, Aim=='SA3' & Habitat_Class=='Supra')
B.3 <- as.data.frame(cbind(with(dat, aggregate(p, list(OTU), mean))[,2], with(dat, aggregate(freq, list(OTU), mean))[,2], with(dat, aggregate(OTU, list(OTU), function(x){as.character(unique(x))}))[,2]))
colnames(B.3) <- c('p.mean', 'freq.mean', 'OTU')
B.3$p.mean <- as.numeric(as.character(B.3$p.mean))
B.3$freq.mean <- as.numeric(as.character(B.3$freq.mean))
Bf.3 <- subset(B.3, OTU %in% tax.focus$OTU)
Bf.3 <- merge(Bf.3, tax.focus, by.x='OTU', by.y='OTU', all=FALSE)
m.fit <- with(B.3, nlsLM(freq.mean ~ pbeta(1/min.seq, min.seq*m*p.mean, min.seq*m*(1-p.mean), lower.tail=FALSE), start=list(m=0.1)))
fun.pred.3 <- function(x){pbeta(1/min.seq, min.seq*coef(m.fit)*x, min.seq*coef(m.fit)*(1-x), lower.tail=FALSE)}

plot.sa3_supra <- ggplot(B.3, aes(x=p.mean, y=freq.mean)) +
	geom_point() +
	geom_point(aes(x=p.mean, y=freq.mean), data=Bf.3, colour='white') +
	geom_point(aes(x=p.mean, y=freq.mean, shape=code), data=Bf.3, colour='red', size=4) +
	scale_shape_manual(values=as.numeric(as.character(Bf.3$code))) +
	stat_function(data=B.3, aes(x=p.mean, y=freq.mean), fun=fun.pred.3, colour='#0072B2') +
	scale_x_log10() +
	ylab("Occurrence frequency") +
	xlab("log(Mean Relative Abundance)") +
	ggtitle("Supragingival communities across Low Flow Cohort") +
	coord_cartesian(ylim=c(-0.025,1.025)) +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),
	   legend.position='none')


#Disease Subgingival
dat <- subset(ncm.preds, Aim=='SA3' & Habitat_Class=='Sub')
B.4 <- as.data.frame(cbind(with(dat, aggregate(p, list(OTU), mean))[,2], with(dat, aggregate(freq, list(OTU), mean))[,2], with(dat, aggregate(OTU, list(OTU), function(x){as.character(unique(x))}))[,2]))
colnames(B.4) <- c('p.mean', 'freq.mean', 'OTU')
B.4$p.mean <- as.numeric(as.character(B.4$p.mean))
B.4$freq.mean <- as.numeric(as.character(B.4$freq.mean))
Bf.4 <- subset(B.4, OTU %in% tax.focus$OTU)
Bf.4 <- merge(Bf.4, tax.focus, by.x='OTU', by.y='OTU', all=FALSE)
m.fit <- with(B.4, nlsLM(freq.mean ~ pbeta(1/min.seq, min.seq*m*p.mean, min.seq*m*(1-p.mean), lower.tail=FALSE), start=list(m=0.1)))
fun.pred.4 <- function(x){pbeta(1/min.seq, min.seq*coef(m.fit)*x, min.seq*coef(m.fit)*(1-x), lower.tail=FALSE)}


plot.sa3_sub <- ggplot(B.4, aes(x=p.mean, y=freq.mean)) +
	geom_point() +
	geom_point(aes(x=p.mean, y=freq.mean), data=Bf.4, colour='white') +
	geom_point(aes(x=p.mean, y=freq.mean, shape=code), data=Bf.4, colour='red', size=4) +
	scale_shape_manual(values=as.numeric(as.character(Bf.4$code))) +
	stat_function(data=B.4, aes(x=p.mean, y=freq.mean), fun=fun.pred.4, colour='#0072B2') +
	scale_x_log10() +
	ylab("Occurrence frequency") +
	xlab("log(Mean Relative Abundance)") +
	ggtitle("Subgingival communities across Low Flow Cohort") +
	coord_cartesian(ylim=c(-0.025,1.025)) +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   axis.line.x = element_line(color="black"),
      axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),
	   legend.position='none')
```

```{r, echo=FALSE, error=FALSE, warning=FALSE, fig.width=24, fig.height=16}
#Actual plot
grid.arrange(plot.sa1_supra, plot.sa1_sub, plot.sa3_supra, plot.sa3_sub, nrow=2)
```

###Figure 3 Bonus: Observed and predicted distribution of microbial taxa for each individual metacommunity
The following are plots just as above but not averaged across individual metacommunities.
```{r, echo=FALSE, error=FALSE, warning=FALSE, results=FALSE, message=FALSE}
tax.focus <- c('SR1_3', 'SR1_2', 'Treponema_sp.1', 'Porphyromonas_gingivalis', 'Tannerella_sp.1', 'Tannerella_forsythia', 'Rothia_dentocariosa', 'Streptococcus_sanguinis', 'Abiotrophia_defectiva', 'Scardovia_wiggsiae', 'Streptococcus_mutans', 'Treponema_denticola')
tax.focus <- data.frame(cbind(tax.focus, seq(from = 97, to = 96+length(tax.focus))))
colnames(tax.focus) <- c('OTU', 'code')

#Define plotting function
sncm.plot <- function(x1){
  dat.f <- subset(x1, OTU %in% tax.focus$OTU)
  dat.f <- merge(dat.f, tax.focus, by.x='OTU', by.y='OTU', all=FALSE)
  m.fit <- with(x1, nlsLM(freq ~ pbeta(1/min.seq, min.seq*m*p, min.seq*m*(1-p), lower.tail=FALSE), start=list(m=0.1)))
  fun.pred.1 <- function(x){pbeta(1/min.seq, min.seq*coef(m.fit)*x, min.seq*coef(m.fit)*(1-x), lower.tail=FALSE)}
  plot.x <- ggplot(x1, aes(x=p, y=freq)) +
    geom_point() +
  	geom_point(aes(x=p, y=freq), data=dat.f, colour='white') +
  	geom_point(aes(x=p, y=freq, shape=code), data=dat.f, colour='red', size=4) +
  	scale_shape_manual(values=as.numeric(as.character(dat.f$code))) +
  	stat_function(data=x1, aes(x=p, y=freq), fun=fun.pred.1, colour='#0072B2') +
  	scale_x_log10() +
  	ylab("Occurrence frequency") +
  	xlab("log(Mean Relative Abundance)") +
  	ggtitle(paste(unique(x1$Subject), unique(x1$Habitat_Class), unique(x1$Tooth_Aspect), sep=' ')) +
  	coord_cartesian(ylim=c(-0.025,1.025)) +
  	theme_bw() +
  	theme(panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     panel.border = element_blank(),
     axis.line.x = element_line(color="black"),
     axis.line.y = element_line(color="black"),
     axis.title.x = element_text(colour='black'),
     axis.text.x = element_text(colour='black'),
     axis.title.y = element_text(colour='black'),
     axis.text.y = element_text(colour='black'),
     legend.position='none')
  return(plot.x)
}

#Compile list of plots
res.list <- list()
for(i in 1:length(unique(ncm.preds$Metacommunity))){
  meta.i <- unique(ncm.preds$Metacommunity)[i]
  dat.i <- subset(ncm.preds, Metacommunity==meta.i)
  res.list[[i]] <- sncm.plot(dat.i)
}
```
```{r, echo=FALSE, error=FALSE, warning=FALSE}
#Plots
for(i in 1:length(res.list)){
  print(res.list[[i]])
}
```

#Part 3: Deviations from the neutral model
We next wanted to investigate whether taxa deviated consistently from the predictions of the model across individuals, occurring either more or less frequently (i.e. being detected in a greater or lesser number of samples) than expected given their average abundance across samples. Taxa that consistently are found more frequently than expected given their average abundance may be under strong positive selection by the host environment or have particularly high dispersal rates, while the oppositie may be true for taxa found less frequently than expected given their average abundance.

For the following two plots, each open point represents the deviation/residuals from the model prediction for each taxa within each subject. Solid red points represent the mean while error bars represent the standard error. There is a dashed vertical line at zero to indicate the model prediction; points to the right of this line represent taxa that are found more frequently than expected, while points to the left of this line represent taxa that are found less frequently than expected.

###Figure X: Deviations from neutral model in SUPRAgingival habitats across subjects
```{r, echo=FALSE, error=FALSE, warning=FALSE, fig.height=42}
dat <- subset(ncm.preds, Habitat_Class=='Supra')
x1 <- with(dat, aggregate(pred.res, list(OTU), mean))[,2]
mres <- as.data.frame(cbind(unique(dat$OTU), x1))
mres$x1 <- as.numeric(as.character(mres$x1))
mres <- mres[with(mres, order(x1)), ]
dat$OTU <- factor(dat$OTU, levels=mres$V1)

ggplot(dat, aes(x=OTU, y=pred.res)) +
	geom_hline(yintercept=0, colour='black', linetype='dashed') +
	geom_point(shape=1, colour='grey50') +
	stat_summary(fun.data = mean_se, geom = "errorbar", position=position_dodge(0.8), width=0.1, colour='#e41a1c') +
	stat_summary(fun.y = mean, geom = "point", position=position_dodge(0.8), colour='#e41a1c') +
	ylab("Deviation from predicted occurrence frequency") +
	xlab("Taxa") +
	coord_flip() +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   strip.background = element_blank(),
	   strip.text.x = element_text(colour='black'),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),,
	   legend.position='none',
	   legend.key = element_blank())
```

###Figure X: Deviations from neutral model in SUBgingival habitats across subjects
```{r, echo=FALSE, error=FALSE, warning=FALSE, fig.height=24}
dat <- subset(ncm.preds, Habitat_Class=='Sub')
x1 <- with(dat, aggregate(pred.res, list(OTU), mean))[,2]
mres <- as.data.frame(cbind(unique(dat$OTU), x1))
mres$x1 <- as.numeric(as.character(mres$x1))
mres <- mres[with(mres, order(x1)), ]
dat$OTU <- factor(dat$OTU, levels=mres$V1)

ggplot(dat, aes(x=OTU, y=pred.res)) +
	geom_hline(yintercept=0, colour='black', linetype='dashed') +
	geom_point(shape=1, colour='grey50') +
	stat_summary(fun.data = mean_se, geom = "errorbar", position=position_dodge(0.8), width=0.1, colour='#e41a1c') +
	stat_summary(fun.y = mean, geom = "point", position=position_dodge(0.8), colour='#e41a1c') +
	ylab("Deviation from predicted occurrence frequency") +
	xlab("Taxa") +
	coord_flip() +
	theme_bw() +
	theme(panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
	   panel.border = element_blank(),
	   strip.background = element_blank(),
	   strip.text.x = element_text(colour='black'),
	   axis.line.x = element_line(color="black"),
       axis.line.y = element_line(color="black"),
	   axis.title.x = element_text(colour='black'),
	   axis.text.x = element_text(colour='black'),
	   axis.title.y = element_text(colour='black'),
	   axis.text.y = element_text(colour='black'),,
	   legend.position='none',
	   legend.key = element_blank())
```


